Certified Knowledge Compilation \\ with Application to Verified Model Counting \\ Experimental Details

### 2023-08-22 doc/JAIR/jair-data.tex
Add experimental data

\section*{Methodology}
\begin{enumerate}
   \item Retrieved 200 public model counting benchmarks from 2022 Model Counting Competition website
     \begin{center}
       \url{https://mccompetition.org/2022/mc_description.html}
     \end{center}
     \begin{itemize}
     \item 100 public instances from Track 1 (unweighted model counting)
     \item 100 public instances from Track 2 (weighted model counting)
     \end{itemize}
   \item Removed duplicates
     \begin{itemize}
    \item  20 instances were duplicated in the two tracks.  
     \item Reduced set of benchmarks to 180 by including some from Track 1 and some from Track 2
     \end{itemize}
   \item Ran D4 on 180 instances with time limit of 4000 seconds
     \begin{itemize}
     \item Completed for 123 instances
     \end{itemize}
   \item Generated POG representations from dDNNF graphs
     \begin{itemize}
     \item Simplifications via constant propagation
       \begin{itemize}
       \item Number of POG nodes around ranged between $0.35\times$
         and $1.0\times$ number of dDNNF nodes (harmonic mean =
         $0.81\times$).  Greater optimizations with smaller
         benchmarks.
       \end{itemize}
     \item Used heuristics to identify root node, and then put in topological order with root node at end.
     \item Successfully generated POG representation for 122 benchmarks
       \begin{itemize}
         \item One failed with too many defining clauses (2,761,457,765).  That count exceeded the 32-bit signed integer used to represent clause identifiers.
       \end{itemize}

     \item Measure complexity by number of defining clauses $D$
       \begin{itemize}
       \item Product or sum node with degree $k$ requires $k+1$ clauses
       \item $D$ also equals sum of number of edges and number of nodes
       \item For 122 instances, Min = 304, Median = 848,784, Mean = 93,107,937, Max = 2,761,457,765
       \end{itemize}

@@

     \item Performance (122 instances):
       \begin{itemize}
       \item Times: Min = 0.012 secs, Max = 2804.5 secs, Median = 3.46 secs
       \item Defining clauses: Min = 304, Max = 1,765,743,261, Median = 774,883
       \item Tree ratio: Min = 1.0, Max = 52,409.7, Median = 11.61, Harmonic Mean = 3.21
       \item Tree Sizes ($D_t$): Min = 304, max = 14,924,670,455,484, median = 7,669,582.
       \end{itemize}
     \end{itemize}
\end{enumerate}

@@

To make the experiments tractable, want to select set of benchmarks that are more tractable.
Limit set to the 92 with at most $10^7$ defining clauses.

D4 Performance (92 instances):
       \begin{itemize}
       \item Times: Min = 0.012 secs, Max = 1672.9 secs, Median = 1.85 secs
       \item Defining clauses: Min = 304, Max = 8,493,275, Median = 378,325
       \item Tree ratio: Min = 1.0, Max = 363.9, Median = 9.37, Harmonic Mean = 2.87
       \item Tree Sizes ($D_t$): Min = 304, max = 1,205,743,897, median = 2,690,776
       \end{itemize}

@@

Our implementation also allows a hybrid of these approaches: at any
point in the traversal, it can generate a proof for a node via a
single call to a SAT solver.  This capability supports both full
monolithic generation (perform monolithic generation at the POG root),
full structured generation (continue traversing to the POG leaves), or
a balance between the two.
The challenge is then to devise a
heuristic criterion for deciding whether to continue the traversal below
some node or to generate a proof for that node monolithically.

Our first set of experiments applies full monolithic and full structured generation to the reduced benchmark set.  We set a time limit of 1000 seconds for the proof generation.

>>>>

\section*{Methodology}
\begin{enumerate}
   \item Retrieved 200 public model counting benchmarks from 2022 Model Counting Competition website
     \begin{center}
       \url{https://mccompetition.org/2022/mc_description.html}
     \end{center}
     \begin{itemize}
     \item 100 public instances from Track 1 (unweighted model counting)
     \item 100 public instances from Track 2 (weighted model counting)
     \end{itemize}
   \item Removed duplicates
     \begin{itemize}
    \item  20 instances were duplicated in the two tracks.  
     \item Reduced set of benchmarks to 180 by including some from Track 1 and some from Track 2
     \end{itemize}
   \item Ran D4 on 180 instances with time limit of 4000 seconds
     \begin{itemize}
     \item Completed for 123 instances
     \end{itemize}
   \item Generated POG representations from dDNNF graphs
     \begin{itemize}
     \item Used heuristics to identify root node, and then put in topological order with root node at end.
     \item Successfully generated POG representation for 122 benchmarks
       \begin{itemize}
         \item One failed with too many defining clauses (2,761,457,765).  That count exceeded the 32-bit signed integer used to represent clause identifiers.
       \end{itemize}

     \item Measure complexity by number of defining clauses $D$
       \begin{itemize}
       \item Product or sum node with degree $k$ requires $k+1$ clauses
       \item $D$ also equals sum of number of edges and number of nodes
       \item For 122 instances, Min = 304, Median = 774,883, Mean = 72,389,514, Max = 1,765,743,261
       \end{itemize}

@@

     \item Performance (122 instances):
       \begin{itemize}
       \item Times: Min = 0.012 secs, Max = 2804.5 secs, Median = 3.46 secs
       \item Defining clauses: Min = 304, Max = 1,765,743,261, Median = 774,883
       \item Tree Sizes ($D_t$): Min = 304, max = 14,924,670,455,484, median = 7,669,582.
       \item Tree ratio: Min = 1.0, Max = 52,409.7, Median = 11.61, Harmonic Mean = 3.21
       \end{itemize}
     \end{itemize}
\end{enumerate}

\section*{Summary Result}

Paper describes two ways to generate the forward implication proof:
structured and monolithic.  Can also combine the two: at any point in
the traversal during structured generation, the forward implication
proof for the entire subgraph can be generated by a single call to a proof-generating
SAT solver.  We'll refer to this as a \emph{hybrid} approach.

Ran some experiments on all but the largest benchmarks (presented
later) seeing under what conditions was it best to use a monolithic,
structured, or hybrid approach, and where the hybrid approach should
shift from structured to monolithic.

General findings:
\begin{itemize}
\item The tree ratio provides a reasonable discriminator for monolithic vs.~structured.  Based on experiments, set threshold at 5.0. Use monolithic below this and structured above.
\item We could not find any way to exploit the hybrid approach for these benchmarks.
\item On the other hand, the monolithic approach becomes impractical beyond around $10^7$ defining clauses.  Even the best SAT solvers become unstable beyond this.
\item Therefore, use hybrid approach for very large problems with tree ratios below 5.0, shift to monolithic once the tree size is at most $10^7$.
\end{itemize}

Our summary data is based on this set of rules.

Summary:

\begin{itemize}
\item Of the 122 benchmarks, 100 were completed using the hybrid approach within 10,000 seconds.
\end{itemize}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 0.70]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
      legend style={at={(0.30,0.96)}},
      legend cell align={left},
                              x post scale=1.8, y post scale=2.4,
                              xmode=log,
                              xmin=0.01,xmax=4000,
                              xtick={0.01, 0.1,1.0,10,100,1000,10000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}, {10,000}},
                              ymin=0.01, ymax=29000,
                              ytick={0.01, 0.1,1.0,10,100,1000,10000,100000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000},{10,000},{100,000}},
                              xlabel={D4 runtime (seconds)}, ylabel={CPOG generation and checking runtime (seconds)},
%                              title={D4 Defining Clause Generation}
            ]
    \input{data-formatted/seconds-summary}
%    \input{data-formatted/seconds-summary-onesided}
    \input{data-formatted/seconds-summary-failures}
    \legend{
      \scriptsize \textsf{Full validation},
%      \scriptsize \textsf{One-sided validation},
      \scriptsize \textsf{Resource limit exceeded},
    }
    \addplot[mark=none, dashed] coordinates{(0.01,25000) (4000, 25000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,20000) (4000,20000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.1,0.01) (10000.0,1000.0)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.01) (10000.0,10000.0)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,0.1) (2000,20000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,1.0) (200, 20000)};
    \addplot[mark=none, color=lightblue] coordinates{(0.01,10.0) (20, 20000)};
    \node[left] at (axis cs: 0.9,0.12) {$0{.}1\times$};
    \node[left] at (axis cs: 0.09,0.12) {$1\times$};
    \node[left] at (axis cs: 0.09,1.2) {$10\times$};
    \node[left] at (axis cs: 0.09,12.0) {$100\times$};
    \node[left] at (axis cs: 0.09,120.0) {$1000\times$};
       \end{axis}

\end{tikzpicture}
} % centering
\caption{Combined runtime for CPOG proof generation and checking  as function of D4 runtime.  Timeouts are shown as points on the dashed line.}
\label{fig:d4:cpog}
\end{figure}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 0.70]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10},
      legend style={at={(0.92,0.28)}},
      legend cell align={left},
                              x post scale=1.9, y post scale=2.10,
                              xmode=log,xmin=100,xmax=1e10, 
                              xtick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000, 1e10}, 
                              xticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$, $10^{10}$},
                              ymode=log, ymin=100, ymax=1800000000, 
                              ytick={100,1000, 10000, 100000, 1000000, 10000000, 100000000, 1000000000},
                              yticklabels={$10^2$,$10^3$,$10^4$,$10^5$,$10^6$,$10^7$,$10^8$,$10^9$},
                              xlabel={Defining Clauses}, ylabel={Proof Clauses},
            ]
    \input{data-formatted/clauses-summary}
%%    \input{data-formatted/clauses-defining-onesided}
    \input{data-formatted/clauses-summary-failures}

    \legend{
      \scriptsize \textsf{Full validation},
%      \scriptsize \textsf{One-sided validation},
      \scriptsize \textsf{No validation},
    }
    \addplot[mark=none, color=lightblue] coordinates{(100,100) (1e9,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1000) (1e8,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,10000) (1e7,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,100000) (1e6,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1000000) (1e5,1e9)};
    \addplot[mark=none, color=lightblue] coordinates{(100,1e9) (1e10,1e9)};
    \addplot[color=black,dashed] coordinates{(1e2,1.5e9) (1e10,1.5e9)};
    \node[left] at (axis cs: 900,1200) {$1\times$};
    \node[left] at (axis cs: 900,12000) {$10\times$};
    \node[left] at (axis cs: 900,120000) {$100\times$};
    \node[left] at (axis cs: 900,1200000) {$1{,}000\times$};
    \node[left] at (axis cs: 900,12000000) {$10{,}000\times$};

          \end{axis}
\end{tikzpicture}
} % centering
\caption{Total number of clauses in CPOG file as function of number of defining clauses}
\label{fig:defining:total}
\end{figure}

@@

To make the experiments tractable, want to select set of benchmarks that are more tractable.
Limit set to the 90 with at most $10^7$ defining clauses, and for which D4 runs in at most 1000 seconds.

D4 Performance (90 instances):
       \begin{itemize}
       \item Times: Min = 0.012 secs, Max = 660.2 secs, Median = 1.72 secs
       \item Defining clauses: Min = 304, Max = 8,493,275, Median = 378,325
       \item Tree Sizes ($D_t$): Min = 304, max = 1,205,743,897, median = 2,690,776
       \item Tree ratio: Min = 1.0, Max = 363.9, Median = 9.98, Harmonic Mean = 2.94
       \end{itemize}

@@

Our implementation also allows a hybrid of these approaches: at any
point in the structured traversal, it can generate a proof for a node via a
single call to a SAT solver.  This capability supports both full
monolithic generation (perform monolithic generation at the POG root),
full structured generation (continue traversing to the POG leaves), or
a balance between the two.
The challenge is then to devise a
heuristic criterion for deciding whether to continue the traversal below
some node or to generate a proof for that node monolithically.

Our first set of experiments applies full monolithic and full structured generation to the reduced benchmark set.  
After some experimenting, we devised the following ``hybrid'' method: when the tree ratio is at most $5.0$, use the monolithic approach, otherwise use the structured approach.
We set a time limit of 1000 seconds for the proof generation and checking.
Overall, our results on the 92 benchmarks were:
\begin{itemize}
\item Monolithic: 83 completed
\item Structured: 82 completed
\item Hybrid: 85 completed
\end{itemize}

\begin{figure}
\centering{%
\begin{tikzpicture}[scale = 0.70]
  \begin{axis}[mark options={scale=0.55},grid=both, grid style={black!10}, ymode=log,
      legend style={at={(0.30,0.96)}},
      legend cell align={left},
                              x post scale=1.8, y post scale=2.0,
                              xmode=log,xmin=0.01,xmax=1000,
                              xtick={0.01, 0.1,1.0,10,100,1000}, xticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
                              ymin=0.01, ymax=2000,
                              ytick={0.01, 0.1,1.0,10,100,1000}, yticklabels={0.01, 0.1, 1.0, 10, 100, {1,000}},
                              xlabel={D4 runtime (seconds)}, ylabel={Toolchain runtime (seconds)},
            ]

    \input{data-formatted/sub10M-seconds-hybrid}
    \legend{
      \scriptsize \textsf{Hybrid},
    }

    \input{data-formatted/sub10M-seconds-failures-hybrid}

          \end{axis}
\end{tikzpicture}
} % centering
\caption{Total runtime for toolchain using three different approaches to forward-implication proof generation, as function of D4 runtime}
\label{fig:d4:cpog:compare}
\end{figure}
